Online Updating of Variational Bayes for Heterogeneous Forecasts of Vehicle Trajectory

### 2018-03-09 Cars/midcanPresentation.tex
presentation

\begin{frame}
\section{Background}
\subsection{Motivation}
\frametitle{Motivation}
\begin{itemize}
\item Bayesian Inference in Forecasting:
\begin{enumerate}
\item Provides forecast densities for the variable of interest.
\item Encorporates parameter uncertainty.
\item Is easy to extend to hierarchical models.
\end{enumerate}
\vspace{3mm}
\item Bayesian Inference is only as good as its compuation:
\begin{enumerate}
\item Computation based on Markov Chain Monte Carlo (MCMC) is 'exact', but slow.
\item Computation based on Variational Bayes (VB) is fast, but approximate.
\end{enumerate}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Motivation}
\begin{itemize}
\item We focus on Hierarchical Models where:
\begin{enumerate}
\item Inference conditioned on existing units in the hierarchy is available.
\item Inference conditioned on additional units in a short time-frame is desired.
\end{enumerate}
\vspace{3mm}
\item We aim to infer, and forecast, additional units in an online setting.
\item We want to leverage the existing information to improve the forecasts.
\end{itemize}
\end{frame}

\begin{frame}
\subsection{Hierarchical Models}
\frametitle{Hierarchical Models}
\begin{itemize}
\item Hierarchical Models are structured into three levels:
\begin{enumerate}
\item The likelihood: $p(z_i | \theta_i)$ for each unit $i$ in the model,
\item The prior: $p(\theta_i | \beta)$ where $\beta$ is shared across all $i$,
\item The hyper-prior: $p(\beta)$ is fixed.
\end{enumerate}
\vspace{5mm}
\item For $i = 1, 2, \dots, N$, we assume that the posterior distribution $$p(\beta, \theta_1, \dots, \theta_N | z_1, \dots, z_N)$$ is available.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Hierarchical Models}
\begin{itemize}
\item What about $i = N+1$? Using $\textbf{z}_{1:j} = \{z_i | i = 1, \dots, j\}$, $$p(\theta_{N+1} |\textbf{z}_{1:N+1}) \propto p(z_{N+1} | \theta_{N+1})\int_{\beta}p(\theta_{N+1} | \beta)p(\beta | \textbf{z}_{1:N})d\beta.$$
\item If there is a time constraint, we might use VB and approximate $$q_{\lambda}(\theta_{N+1} | \textbf{z}_{1:N+1}) \approx p(\theta_{N+1} |\textbf{z}_{1:N+1})$$
\item choosing $$\lambda_{VB} = \arg \underset{\lambda}{\min}\mbox{ }D(p, q).$$
\end{itemize}
\end{frame}

\begin{frame}
\subsection{Variational Bayes}
\frametitle{Variational Bayes and the Kullback-Leibler Divergence}
\begin{itemize}
\item A common divergence function is the Kullback Leibler Divergence from $q$ to $p$,
\begin{align}
\label{KL-def}
KL[q_{\lambda}(\theta_{N+1} &| \textbf{z}_{1:N+1})\hspace{.1cm}||\hspace{.1cm}p(\theta_{N+1} | \textbf{z}_{1:N+1})] = \nonumber \\
&E_q \left[ \log(q_{\lambda}(\theta_{N+1} | \textbf{z}_{1:N+1})) - \log(p(\theta_{N+1} | \textbf{z}_{1:N+1})) \right],
\end{align}
\item The expectation is intractable for most $p$ and $q$.
\item It typically cannot be estimated with Monte-Carlo simulations.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{The Evidence Lower Bound}
\begin{itemize}
\item Instead look at the Evidence Lower Bound (ELBO), 
\begin{equation}
\label{ELBO}
\mathcal{L}(q, \lambda) = E_{q} \left[\log(p(\theta_{N+1}, \textbf{z}_{1:N+1})) - \log(q_{\lambda}(\theta | \textbf{z}_{1:N+1}))\right],
\end{equation}
\item Stochastic Estimates are available through Monte-Carlo: $$\mathcal{L}(q, \lambda) \approx \frac{1}{M} \sum_{j=1}^M \left(\log(p(\theta_{j, N+1}, \textbf{z}_{1:N+1})) - \log(q_{\lambda}(\theta_{j, N+1} | \textbf{z}_{1:N+1})) \right)$$ where $\theta_{j, N+1} \sim q_{\lambda}(\theta_{N+1} | \textbf{z}_{1:N+1})$.
\item Maximising the ELBO is equivalent to minimising the Kullback-Leibler Divergence.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Stochastic Optimisation}
\begin{itemize}
\item For an exponential family $p$ and factorisable $q$ Mean Field Variational Bayes provides a fast method to the ELBO. 
\item We do not restrict $p$ and $q$ this way and instead use Stochastic Gradient Ascent.
\item Repeat until $| \mathcal{L}(q, \lambda^{(m)}) - \mathcal{L}(q, \lambda{(m+1)})| < \nu$:
\begin{equation}
\label{gradientAscent}
\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)} \widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}} \bigg\rvert_{\lambda = \lambda^{(m)}}
\end{equation}
\end{itemize}
\begin{figure}
\centering
\includegraphics[width = 0.5\textwidth]{gradientascent}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Stochastic Optimisation}
\begin{itemize}
\item ELBO gradients can be estimated by the Score Estimator:
\begin{equation}
\label{scoreDeriv}
\widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}}_{SC} = \sum_{j = 1}^M \frac{\delta \log(q_{\lambda}(\theta_{j, i} | \textbf{z}))}{\delta \lambda} \left(\log(p(\theta_{j, i}, \textbf{z})) - \log(q_{\lambda}(\theta_{j, i} | \textbf{z})) \right),
\end{equation}
where $\theta_j \sim q_{\lambda}(\theta | \epsilon)$.
\item Or the Reparametersied Estimator, where $\theta = f(\epsilon, \lambda)$:
\begin{equation}
\label{rpDeriv}
\widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}}_{RP} = \sum_{j = 1}^M \frac{\delta \theta}{\delta \lambda} \frac{\delta \log(p(\theta, \textbf{z}))}{\delta \theta} \bigg\rvert_{\theta = f(\lambda, \epsilon_j)} + \frac{\delta J(\lambda, \epsilon_j)}{\delta \lambda}, 
\end{equation}
where $\epsilon_j \sim q(\epsilon)$.
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Updating Gradient Estimators}
\begin{itemize}
\item The updating score estimator:
\begin{align}
\widehat{\frac{\delta\mathcal{L}(q, \lambda_T)}{\delta \lambda_T}}_{USC} &= \sum_{j = 1}^M \frac{\delta \log(q_{\lambda_T}(\theta_{i, j} | z_{i, 1:T}))}{\delta \lambda_T} \nonumber \\
&\times \left(\log(q_{\lambda_S}(\theta_{i, j} | z_{i, 1:S}) - \log(q_{\lambda_T}(\theta_{i, j} | z_{i, 1:T})) \right) \label{scoreUpdate}
\end{align}
where $\theta_{i, j} \sim q(\theta_{i} | z_{i, 1:T})$. 
\item The updating reparameterised estimator:
\begin{equation}
\label{rpUpdate}
\widehat{\frac{\delta\mathcal{L}(q, \lambda_T)}{\delta \lambda_T}}_{URP} = \sum_{j = 1}^M \frac{\delta f(\lambda_T, \epsilon_j)}{\delta \lambda_T} \frac{\delta \log(q_{\lambda_S}(\theta_{i} |z_{i, 1:S}))}{\delta \theta_{i}} \bigg\rvert_{\theta_{i} = f(\lambda_T, \epsilon_j)} + \frac{\delta J(\lambda_T, \epsilon_j)}{\delta \lambda_T},
\end{equation}
where $\epsilon_j \sim q(\epsilon)$.
\end{itemize}
\end{frame}

>>>>

\begin{frame}
\section{Background}
\subsection{Motivation}
\frametitle{Motivation}
\begin{itemize}
\item We aim to build a framework to provide density forecasts for new units in a heirarchy as they are observed.
\pause
\item Bayesian Inference in Forecasting:
\begin{enumerate}
\item Provides density forecasts for the variable of interest.
\item Provides a mechanism to update posterior distributions
\item Is easy to extend to hierarchical models.
\end{enumerate}
\pause
\vspace{3mm}
\item Bayesian Inference is only as good as its compuation:
\begin{enumerate}
\item Computation based on Markov Chain Monte Carlo (MCMC) is 'exact', but slow.
\item Computation based on Variational Bayes (VB) is fast, but approximate.
\end{enumerate}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Motivation}
\begin{itemize}
\item We focus on Hierarchical Models where:
\begin{enumerate}
\item Inference conditioned on existing units in the hierarchy is available.
\item Inference conditioned on additional units in a short time-frame is desired.
\end{enumerate}
\vspace{3mm}
\pause
\item We aim to infer, and forecast, additional units in an online data setting.
\pause
\item We want to leverage the existing information to improve the forecasts.
\end{itemize}
\end{frame}

\begin{frame}
\subsection{Hierarchical Models}
\frametitle{Hierarchical Models}
\begin{itemize}
\item Hierarchical Models are structured into three levels:
\begin{enumerate}
\item The likelihood: $p(z_i | \theta_i)$ for each unit $i$ in the model,
\item The prior: $p(\theta_i | \beta)$ where $\beta$ is shared across all $i$,
\item The hyper-prior: $p(\beta)$ is fixed.
\end{enumerate}
\pause
\vspace{5mm}
\item For $i = 1, 2, \dots, N$, we assume that the posterior distribution $$p(\beta, \theta_1, \dots, \theta_N | z_1, \dots, z_N)$$ is available.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Hierarchical Models}
\begin{itemize}
\item What about $i = N+1$? Using $\textbf{z}_{1:j} = \{z_i | i = 1, \dots, j\}$, $$p(\theta_{N+1} |\textbf{z}_{1:N+1}) \propto p(z_{N+1} | \theta_{N+1})\int_{\beta}p(\theta_{N+1} | \beta)p(\beta | \textbf{z}_{1:N})d\beta.$$
\pause
\item If there is a time constraint, we might use VB with the approximation $$q_{\lambda}(\theta_{N+1} | \textbf{z}_{1:N+1}) \approx p(\theta_{N+1} |\textbf{z}_{1:N+1})$$
\item choosing $$\lambda_{VB} = \arg \underset{\lambda}{\min}\mbox{ }D(p, q).$$
\end{itemize}
\end{frame}

\begin{frame}
\subsection{Variational Bayes}
\frametitle{Variational Bayes and the Kullback-Leibler Divergence}
\begin{itemize}
\item A common divergence function is the Kullback Leibler Divergence from $q$ to $p$,
\begin{align*}
\label{KL-def}
KL[q_{\lambda}(\theta_{N+1} &| \textbf{z}_{1:N+1})\hspace{.1cm}||\hspace{.1cm}p(\theta_{N+1} | \textbf{z}_{1:N+1})] = \\
&E_q \left[ \log(q_{\lambda}(\theta_{N+1} | \textbf{z}_{1:N+1})) - \log(p(\theta_{N+1} | \textbf{z}_{1:N+1})) \right],
\end{align*}
\pause
\item The expectation is intractable for most $p$ and $q$.
\item It typically cannot be estimated with Monte-Carlo simulations.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{The Evidence Lower Bound}
\begin{itemize}
\item Instead look at the Evidence Lower Bound (ELBO), $$\mathcal{L}(q, \lambda) = E_{q} \left[\log(p(\theta_{N+1}, \textbf{z}_{1:N+1})) - \log(q_{\lambda}(\theta | \textbf{z}_{1:N+1}))\right],$$
\pause
\item Stochastic Estimates are available through Monte-Carlo: $$\mathcal{L}(q, \lambda) \approx \frac{1}{M} \sum_{j=1}^M \left(\log(p(\theta_{N+1}^{(j)}, \textbf{z}_{1:N+1})) - \log(q_{\lambda}(\theta_{N+1}^{(j)} | \textbf{z}_{1:N+1})) \right)$$ where $\theta_{N+1}^{(j)} \sim q_{\lambda}(\theta_{N+1} | \textbf{z}_{1:N+1})$.
\pause
\item Maximising the ELBO is equivalent to minimising the Kullback-Leibler Divergence.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Stochastic Optimisation}
\begin{itemize}
\item For an exponential family $p$ and factorisable $q$ Mean Field Variational Bayes provides a fast method to the ELBO. 
\item We do not restrict $p$ and $q$ this way and instead use Stochastic Gradient Ascent.
\item Repeat until $| \mathcal{L}(q, \lambda^{(m)}) - \mathcal{L}(q, \lambda{(m+1)})| < \nu$:
\begin{equation}
\label{gradientAscent}
\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)} \widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}} \bigg\rvert_{\lambda = \lambda^{(m)}}
\end{equation}
\end{itemize}
\begin{figure}
\centering
\includegraphics[width = 0.4\textwidth]{gradientascent}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Stochastic Optimisation}
\begin{itemize}
\item ELBO gradients can be estimated by the Score Estimator:
\begin{align*}
\widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}}_{SC} &= \sum_{j = 1}^M \frac{\delta \log(q_{\lambda}(\theta_{N+1}^{(j)} | \textbf{z}_{1:N+1}))}{\delta \lambda}  \\
&\times \left(\log(p(\theta_{N+1}^{(j)}, \textbf{z}_{1:N+1})) - \log(q_{\lambda}(\theta_{N+1}^{(j)} | \textbf{z}_{1:N+1})) \right),
\end{align*}
where $\theta_{N+1}^{(j)} \sim q_{\lambda}(\theta_{N+1} | \textbf{z}_{1:N+1})$.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Stochastic Optimisation}
\begin{itemize}
\item Or the Reparameterised Estimator, where $\theta_{N+1} = f(\epsilon, \lambda)$:
\begin{align*}
\widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}}_{RP} &= \sum_{j = 1}^M \frac{\delta J(\lambda, \epsilon^{(j)})}{\delta \lambda} + \frac{\delta \theta_{N+1}}{\delta \lambda} \\
\times \frac{\delta \log(p(\theta_{N+1}, \textbf{z}_{1:N+1}))}{\delta \theta_{N+1}} \bigg\rvert_{\theta_{N+1} = f(\lambda, \epsilon^{(j)})},
\end{align*}
where $\epsilon^{(j)} \sim q(\epsilon)$.
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Updating Gradient Estimators}
\begin{itemize}
\item The updating score estimator:
\begin{align*}
\widehat{\frac{\delta\mathcal{L}(q, \lambda_T)}{\delta \lambda_T}}_{USC} &= \sum_{j = 1}^M \frac{\delta \log(q_{\lambda_T}(\theta_{N+1}^{(j)} | z_{N+1, 1:T}))}{\delta \lambda_T}  \\
&\times \left(\log(q_{\lambda_S}(\theta_{N+1}^{(j)} | z_{N+1, 1:S}) - \log(q_{\lambda_T}(\theta_{N+1}^{(j)} | z_{N+1, 1:T})) \right)
\end{align*}
where $\theta_{N+1}^{(j)} \sim q(\theta_{N+1} | z_{N+1, 1:T})$. 
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Updating Gradient Estimators}
\begin{itemize}
\item The updating reparameterised estimator:
\begin{align*}
\widehat{\frac{\delta\mathcal{L}(q, \lambda_T)}{\delta \lambda_T}}_{URP} &= \sum_{j = 1}^M \frac{\delta J(\lambda_T, \epsilon^{(j)})}{\delta \lambda_T} + \frac{\delta f(\lambda_T, \epsilon^{(j)})}{\delta \lambda_T} \\
&\times \frac{\delta \log(q_{\lambda_S}(\theta_{N+1} |z_{N+1, 1:S}))}{\delta \theta_{N+1}} \bigg\rvert_{\theta_{N+1} = f(\lambda_T, \epsilon^{(j)}} 
\end{align*}
where $\epsilon^{(j)} \sim q(\epsilon)$.
\end{itemize}
\end{frame}
